### 2018-03-15 paper/3-abstractions.tex
Finalise S3

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot even
  use the application operator \hs{<*>}, which limits dependencies to a single
  linear chain.
  % It is interesting to note that if \hs{busy} is given a \emph{functorial task}
  % of type \hs{Task}~\hs{Functor}, which cannot use the application operator
  % \hs{<*>}, then \hs{busy} can rejoice in knowing that it will never have to
  % rebuild a key twice. Indeed, functorial tasks are very much like tail recursive
  % functions, which cannot branch into multiple subproblems.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
argument, an application of \hs{busy} to a \hs{Task}~\hs{Applicative} will
typecheck just fine.  It feels a bit like sub-typing, but it is actually just
ordinary higher-rank polymorphism at work.

@@

To express this formally we need an auxiliary function \hs{compute}, that
computes the value of a key in a given store \emph{without attempting to update
any dependencies}:
\begin{minted}[xleftmargin=10pt]{haskell}
recompute :: Task Monad k v -> Store i k v -> k -> Maybe v
recompute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}

Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the pure
function \hs{getValue}~\hs{store}, whose type is \store, and the \hs{fetch}
argument of the \hs{task}, whose type must be \hs{k}~\hs{->}~\hs{f}~\hs{v} for
some \hs{f}. To fix the mismatch, we wrap the result of the pure function in the
\hs{Identity} monad: the function \hs{Identity}~\hs{.}~\hs{getValue}~\hs{store}
has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
contents of \hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = build task key store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of recomputing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} is provided in the Appendix, which
    is dedicated to the task dependency analysis. This requirement guarantees
    that (i) no inputs were corrupted during the build, and (ii)~the \hs{ideal}
    store has constraints not only on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task k = case task (\k -> Const [k]) k of
                         Nothing         -> []
                         Just (Const ks) -> ks
\end{minted}
\vspace{1mm}
Here \hs{Const} is a standard Haskell type defined in Figure~\ref{fig-types}.
We instantiate \hs{f} to \hs{Const [k]}.  So a value of type \hs{f v},
or in this case \hs{Const [k] v}, contains no value \hs{v}, but does contain
a list of keys of type \hs{[k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task}
records a single dependency; and the standard definitions of \hs{Applicative}
for \hs{Const} (which we give in Figure~\ref{fig-types}) combines the
dependencies from different parts of the task.
As a result, running the task with \hs{f} = \hs{Const [k]} will accumulate a list of
the task's dependencies -- and that is just what \hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
Notice that these calls to \hs{dependencies} do no actual computation
(in this case, spreadsheet arithmetic). They can't:
we are not supplying a store or any input numbers.
So, through the wonders of polymorphism, we are able to extract the
dependencies of the spreadsheet formula, and to do so efficiently,
simply by running its code in a different \hs{Applicative}!
This is not new -- e.g. see~\cite{free-applicatives} -- but it is cool.

So much for applicatives.  What about monadic tasks with dynamic dependencies?
We introduce the function \hs{track}:
a combination of \hs{recompute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
track task store = fmap runWriter . task (\k -> writer (getValue store k, [k]))
\end{minted}
\vspace{1mm}

The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
we feed \hs{fetch k = writer (getValue store k, [k])} to the task, which in
addition to fetching a value from the store, tracks the corresponding key. We
can \hs{track} both applicative and monadic tasks:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ track sprsh1 store "B1"
Just (30,["A1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track sprsh2 store "B1"
Just (20,["C1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store' = initialise () (\key -> if key == "A1" then 10 else 1)
@\ghci@ track sprsh2 store' "B1"
Just (10,["C1","A1"])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of \hs{sprsh2} are determined by the contents of
the store.
\todo{slpj}{This track stuff is not well motivated. When do we want it?  How do we use it?}

\todo{slpj}{I think we should push 'inputs' into the appendix.}

Here is a draft implementation of \hs{inputs} used in the definition of
build system correctness in \S\ref{sec-build-correctness}:

\begin{minted}[xleftmargin=10pt]{haskell}
inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
inputs task store key = filter (isInput task) (closure deps key)
  where
    deps k = maybe [] snd (track task store k)

closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure

data Proxy a = Proxy

isInput :: Task Monad k v -> k -> Bool
isInput task = isNothing . task (const Proxy)
\end{minted}

\subsection{Free tasks (?)}

\todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
intuitions on what tasks really are. Requires some further discussion with Neil.
This will take at least one page.}

% \subsection{Recognising an applicative compute wearing a monadic hat}

% Although dynamic dependencies are very useful, they typically represent only a
% small fraction of all dependencies. One can exploit this in a build system by
% recognising applicative computations even though they have the type
% \hs{Compute}~\hs{Monad} and treating them specially.

% Consider the following data type.

% \subsection{Free build structures}\label{sec-free-build}

% \begin{minted}{haskell}
% data Depend k v r = Depend [k] ([v] -> r)

% data Depends k v r = Depends [k] ([v] -> Depends k v r)
%                    | Done r
% \end{minted}

>>>>

So the type \hs{Task} emerges naturally, almost inevitably.  But now that
it \emph{has} emgerged, we find the that constraints \hs{c} classify task descriptions
in a very interesting way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic task
  description allows \emph{dynamic} dependencies, in which a formula may depend
  on the value of cell \cmd{C}, but \emph{which} cell \cmd{C} depends on the
  value of another cell \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: the task description cannot
  even use the application operator \hs{<*>}, which limits dependencies to a
  single linear chain. It is interesting to note that, when run on a
  \hs{Task}~\hs{Functor}, the \hs{busy} build system will build each key at most
  once, thus partially fulfilling the minimality requirement~\ref{def-minimal}.
  Alas, it still has no mechanism to decide which input keys changed since the
  previous build.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\end{itemize}

Notice also that, even though \hs{busy} takes a \hs{Task}~\hs{Monad} as its
argument, an application of \hs{busy} to a \hs{Task}~\hs{Functor} or
a \hs{Task}~\hs{Applicative} will typecheck just fine. It feels a bit like
sub-typing, but it is actually just ordinary higher-rank polymorphism at
work~\cite{jones2007practical}.

@@

To express this formally we need an auxiliary function \hs{compute}, that
computes the value of a key in a given store \emph{without attempting to update
any dependencies}:
\begin{minted}[xleftmargin=10pt]{haskell}
compute :: Task Monad k v -> Store i k v -> k -> Maybe v
compute task store key = runIdentity <$> task (Identity . getValue store) key
\end{minted}

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-types}).
The use of \hs{Identity} just fixes the `impedance mismatch' between the pure
function \hs{getValue}~\hs{store}, whose type is \store, and the \hs{fetch}
argument of the \hs{task}, whose type must be \hs{k}~\hs{->}~\hs{f}~\hs{v} for
some \hs{f}. To fix the mismatch, we wrap the result of the pure function in the
\hs{Identity} monad: the function \hs{Identity}~\hs{.}~\hs{getValue}~\hs{store}
has the type \hs{k}~\hs{->}~\hs{Identity}~\hs{v}, and can now be passed to a
\hs{task}. The result comes as \hs{Maybe}~\hs{(Identity}~\hs{v)}, hence we now
need to get rid of the \hs{Identity} wrapper by applying \hs{runIdentity} to the
contents of \hs{Maybe}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
build         :: Build c i k v
task          :: Task c k v
key           :: k
store, @@result :: Store i k v
result = @@build @@task @@key @@store
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of recomputing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
    \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
    \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
    \[
    \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
    \]
    % where the implementation of \hs{inputs} is provided in the Appendix, which
    % is dedicated to the task dependency analysis.
    This requirement asserts that (i) no inputs were corrupted during the
    build, and (ii)~the \hs{ideal} store has constraints not only on the output,
    but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

Earlier we remarked that a \hs{Task}~\hs{Applicative} could only have static
dependencies. Usually we would extract such static dependencies by (in the case
of \Excel) looking at the syntax tree of the formula.  But a task description
has no such syntax tree. Yet, remarkably, we can use the polymorphism of a
\hs{Task}~\hs{Applicative} to find its dependencies \emph{without doing any of
the actual work}. Here is the code:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Task Applicative k v -> k -> [k]
dependencies task key = case task (\k -> Const [k]) key of
                            Nothing         -> []
                            Just (Const ks) -> ks
\end{minted}
\vspace{1mm}
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-types}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}.  So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-types}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is just what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
Notice that these calls to \hs{dependencies} do no actual computation (in this
case, spreadsheet arithmetic). They cannot: we are not supplying a store or any
input numbers. So, through the wonders of polymorphism, we are able to extract
the dependencies of the spreadsheet formula, and to do so efficiently, simply by
running its code in a different \hs{Applicative}! This is not new, for example
see~\cite{free-applicatives}, but it is cool.

So much for applicative tasks. What about monadic tasks with dynamic
dependencies? As we have seen in~\S\ref{sec-background-shake}, dynamic
dependencies need to be tracked too. This cannot be done statically; note that
the application of the function \hs{dependencies} to a \hs{Task}~\hs{Monad} will
not typecheck. We need to run a monadic task on a store with concrete values,
which will determine the discovered dependencies. Accordingly, we introduce
the function \hs{track}: a combination of \hs{compute} and \hs{dependencies}
that computes both the resulting value and the list of its dependencies:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
track task store = runWriter . task (\k -> writer (getValue store k, [k]))
\end{minted}
\vspace{1mm}

\noindent
The implementation uses the standard Haskell \hs{Writer} monad: we instantiate
\hs{f} to \hs{Writer}~\hs{[@@k]}~\hs{v}, and feed
\hs{fetch}~\hs{k}~\hs{=}~\hs{writer}~\hs{(@@getValue}~\hs{store}~\hs{k,}~\hs{[@@k])}
to the task, which in addition to fetching a value from the store, tracks the
corresponding key. Here is an example of \hs{track}ing monadic tasks:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store1 = initialise () (\key -> if key == "A1" then 10 else 1)
@\ghci@ track sprsh2 store1 "B1"
Just (1,["C1","B2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store2 = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ track sprsh2 store2 "B2"
Just (20,["C1","B1"])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of \hs{sprsh2} (see the spreadsheet
in~\S\ref{sec-task-monad}) are determined by the contents of the store.
Furthermore, the above example highlights the static cycle between cells
\cmd{B1} and \cmd{B2}: in \hs{store1}, we have \cmd{B1 = B2 = A1 = 10},
whereas in \hs{store2}, we have \cmd{B2 = B1 = A2 = 20}.

% \subsection{Free tasks}
% \todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
% intuitions on what tasks really are. This will take at least one page.}
%
% data TaskA k v a = InputA | TaskA [k] ([v] -> a)
%
% data TaskM k v a = InputM | DoneM a | TaskM [k] ([v] -> Task k v a)

% \subsection{Recognising an applicative task wearing a monadic hat}
% Although dynamic dependencies are very useful, they typically represent only a
% small fraction of all dependencies. One can exploit this in a build system by
% recognising applicative tasks even though they have the type
% \hs{Task}~\hs{Monad} and treating them specially.

### 2018-03-15 paper/9-appendix.tex
Add description of track
Finalise S3

\todo{AM}{Explain \hs{trackDependencies}.}

>>>>

\todo{AM}{Explain \hs{track}.}

Here is a draft implementation of \hs{inputs} used in the definition of
build system correctness in \S\ref{sec-build-correctness}:

\begin{minted}[xleftmargin=10pt]{haskell}
inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
inputs task store key = filter (isInput task) (closure deps key)
  where
    deps k = maybe [] snd (track task store k)

closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure

data Proxy a = Proxy

isInput :: Task Monad k v -> k -> Bool
isInput task = isNothing . task (const Proxy)
\end{minted}
